 ---
 title: "Practical 2: Electronic Medical Records and NLP"
 author: "sarah de Lima"
 date: "2026-06-05"
 output: html_document
 ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.width  = 9,
  fig.height = 5
)

set.seed(42)  # global seed for reproducibility

library(here)        # reproducible file paths
library(tidyverse)   # dplyr, ggplot2, stringr, tidyr, readr, forcats
library(tidytext)    # tidy text mining
library(SnowballC)   # Porter stemmer
library(textstem)    # lemmatization
library(clinspacy)   # MTSamples dataset
library(topicmodels) # Latent Dirichlet Allocation
library(scales)      # axis label helpers
library(sessioninfo) # session reproducibility report
library(reshape2)    # data reshaping library
```

If the above code-block errors, install any missing libraries (`install.packages(c('LIBRARYNAME1', 'LIBRARYNAME2'))`)

## Introduction

This practical explores natural language processing (NLP) of unstructured clinical
free-text from electronic medical records (EMR). You will work through an analysis going from raw
text to an unsupervised topic model using the `MTSamples` medical transcription dataset.

The analyses you will do mirror some of those taken in the assigned readings and refer back to them for some questions:

- **Sun et al. (2024)** (Latent Dirichlet Allocation (LDA)-based topic modelling for clinical social work notes)
- **Chung et al. (2024)** (GPT-4 Turbo for classifying perioperative outcomes from pre-operative notes).

**Learning objectives:** by the end of this practical you should understand how to:

- apply regular expressions to search clinical free-text
- tokenise, normalise, stem, and lemmatize medical text
- compute TF-IDF to identify specialty-discriminating terms
- build a document-term matrix and fit an LDA topic model on clinical notes
- compare traditional NLP with LLM-based approaches

---

## Data Loading and Exploration

This lab is going to use the `MTSamples` medical transcription dataset.  You can find the original source of this data [here](https://www.mtsamples.com/) (ignore any HTTP vs HTTPS related warnings).

```{r load-data}
raw_data <- dataset_mtsamples()
glimpse(raw_data)
```

I've not supplied a data dictionary with this dataset. This is a VERY common situation with real EMR data.

**Q1a (2 marks):** Based on the `glimpse()` output create a simple data dictionary (variable name with a 1-2 sentence explanation of the variable).
```{r}
glimpse(raw_data)
```
note_id = Identification number for the electronic medical note. 
description = clinician notes during assessment
medical specialty = department or specialty where assessment took place
sample name = Procedure administered
transcription = medical history and detailed description of interaction
keywords = words that reflect the nature of the visit pulled from the transcription.

**Q1b (2 marks):** What sort of information are you missing about how this dataset was created that could be important to your analyses?

I would like to know if this data is independent, or if the entries may be coming from the same patient at different times.  

**Q1c (2 marks):** How many distinct medical specialties are represented?  What are the most common 5? Create a ggplot bar graph showing how many transcripts come from each specialty.  

40 distinct specialties are represented. The most common 5 are Surgery, Consult history, cardiovascular and pulmonary, orthopedic, and Radiology.

```{r specialty-bar}
spec <- unique(raw_data$medical_specialty) %>%
  length()
  
raw_data %>%
  ggplot(aes(x = medical_specialty )) + 
  geom_bar(fill = "deeppink4") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Medical Specialty", y = "number of charts", title = "Number of chart records by specialty type")
```


To make patterns slightly easier to interpret for the sake of learning, we will simplify our dataset down to just transcripts from three specialties that have relatively distinct clinical vocabularies: neurological medicine, musculoskeletal/orthopaedic, and diagnostic imaging. In real analyses this is something you would have to infer from the data as part of your exploratory data analysis.

```{r filter-data}
filtered_data <- raw_data |>
  filter(medical_specialty %in% c("Orthopedic", "Radiology", "Neurology"))

count(filtered_data, medical_specialty)
```

---

## Regular Expressions

Now we are going to learn how to use the `stringr` library for regular expressions in R.
Regular expressions (regex) enable flexible pattern-based searches over text. 
As discussed in the lecture, they are widely used in NLP (including as part of the framework around LLM systems) for
finding measurements, medications, and semi- structured values embedded in free-text.

Key `stringr` functions:

| Function | Purpose |
|---|---|
| `str_detect(x, pat)` | Returns `TRUE`/`FALSE` per string |
| `str_count(x, pat)` | Number of non-overlapping matches |
| `str_extract(x, pat)` | First match |
| `str_extract_all(x, pat)` | All matches (returns a list) |


Key regex special characters (in R strings, backslashes must be doubled — e.g., `\d` is written `\\d`):

| Pattern | Meaning | Example match |
|---|---|---|
| `\\d` | Any digit `[0-9]` | `"120"` in `"120 mg"` |
| `\\D` | Any non-digit | `"mg"` in `"120 mg"` |
| `\\w` | Word character `[a-zA-Z0-9_]` | `"bp"` in `"bp_value"` |
| `\\W` | Non-word character | `" "` in `"120 mg"` |
| `\\s` | Whitespace (space, tab, newline) | `" "` in `"120 mg"` |
| `\\S` | Non-whitespace | `"120"` in `"120 mg"` |
| `\\b` | Word boundary | matches position between `"120"` and `" "` |
| `.` | Any character except newline | `"1"`, `"a"`, `"/"` |
| `^` | Start of string | `"BP"` in `"BP: 120/80"` |
| `$` | End of string | `"mmHg"` at end of `"120/80 mmHg"` |
| `*` | 0 or more of preceding | `"\\d*"` matches `""` or `"123"` |
| `+` | 1 or more of preceding | `"\\d+"` matches `"1"` or `"123"` |
| `?` | 0 or 1 of preceding (optional) | `"mmHg?"` matches `"mmH"` or `"mmHg"` |
| `{n}` | Exactly *n* of preceding | `"\\d{3}"` matches `"120"` |
| `{n,m}` | Between *n* and *m* of preceding | `"\\d{2,3}"` matches `"12"` or `"120"` |
| `[abc]` | Any one character in set | `"[aeiou]"` matches any vowel |
| `[^abc]` | Any character *not* in set | `"[^0-9]"` matches non-digits |
| `(a\|b)` | Either `a` or `b` | `"(mg\|mcg)"` matches `"mg"` or `"mcg"` |
| `( )` | Capture group | `"(\\d+) mg"` captures the number |
| `[:alpha:]` | Any letter (POSIX class) | equivalent to `[a-zA-Z]` |
| `[:digit:]` | Any digit (POSIX class) | equivalent to `\\d` |
| `[:alnum:]` | Any letter or digit | equivalent to `[a-zA-Z0-9]` |
| `[:space:]` | Any whitespace (POSIX class) | equivalent to `\\s` |
| `[:punct:]` | Any punctuation character | `"."`, `","`, `"!"` |

### Finding blood pressure measurements

Blood pressure is typically recorded as systolic/diastolic (e.g., "120/80" or
"130/85 mmHg").  We can therefore look for a pair of 2-3 digit numbers (`\\d{2,3}`) separated by a `/` (i.e., `\\d{2,3}/\\d{2,3}`) optionally followed by a pressure-related `mmHg` unit (with or without a preceding space `\\s*`). The special regex characters like `\\d` and `\\s` have that extra `\` so they are "escaped" and not treated as a literal character.

```{r regex-bp}
bp_pattern <- "\\d{2,3}/\\d{2,3}(\\s*mmHg)?"


filtered_data |>
  mutate(has_bp = str_detect(transcription, bp_pattern)) |>
  count(medical_specialty, has_bp) |>
  pivot_wider(names_from = has_bp, values_from = n,
              names_prefix = "bp_") |>
  mutate(pct_with_bp = percent(bp_TRUE / (bp_TRUE + bp_FALSE),
                                       accuracy = 0.1))
```

### Counting medication dosages

Now let's look for medication doses.  We can make a regex that finds one of the standard dosage units (e.g., `mg` `mcg`, `ml`, `mL` `unit`) preceded by some sort of number (`\\d+`) with or without a decimal point (`\\.?\\d*`).

```{r regex-dose}
dose_pattern <- "\\d+\\.?\\d*\\s*(mg|mcg|ml|mL|units?)"

filtered_data |>
    mutate(n_doses = str_count(transcription, dose_pattern)) |>
    filter(n_doses > 0) |>
    group_by(medical_specialty) |>
    summarise(mean_doses   = round(mean(n_doses), 1),
              median_doses = median(n_doses))
```

**Q2 (3 marks):** Write a regular expression to find temperature measurements
in clinical text (e.g., "98.6 F", "37.2°C", "temperature of 38.1", "afebrile").
Report how many notes in each specialty contain at least one temperature mention. 

Regex builders like [this website](https://spannbaueradam.shinyapps.io/r_regex_tester/) can make it a lot quicker to iterate and improve your regular expressions!

133 notes from Neurology, 120 notes from Orthopedic, and 82 notes from Radiology make at least one mention of temperature.

```{r q2}
#"\\d+\\.?\\d*\\d*\\°?\\s*(F|C)"

temp_pattern <- "(\\d+\\.?\\d*\\°?\\s*(F|C))|(afebrile)|(temperature)|(Temperature)|(Temp)|(temp)|(Temp:)|(febrile)"


filtered_data |>
    mutate(n_temp = str_count(transcription, temp_pattern)) |>
    filter(n_temp > 0) |>
    group_by(medical_specialty) |>
    summarise(n_temp   = round(length(n_temp), 1))

```



---

## Text Normalisation

A key stage in traditional text analysis is to normalise raw text — reducing variability due to capitalisation, punctuation, and common function words — before any modelling or analysis.

### Tokenisation and stop word removal

`tidytext::unnest_tokens()` splits text into individual tokens and lowercases
everything. We then strip residual punctuation, drop very short tokens, and
remove stop words (common words like "the", "of", "and" that carry little
domain information):

```{r tokenise}
tokenised_data <- filtered_data |>
  unnest_tokens(word, transcription, to_lower = TRUE) |>
  mutate(word = str_replace_all(word, "[^[:alpha:]]", "")) |>
  filter(str_length(word) > 1) |>
  anti_join(stop_words, by = "word")

glimpse(tokenised_data)
```

**Q3 (2 marks):** How many stop words are in the `stop_words` list, how many different `lexicon` are included, and which `lexicon` contains the most stop words? Answer these questions using code below and create a bargraph of the number of unique stop words per lexicon.

There are 1149 stop words, and 3 lexicons. SMART has the most unique words (571 unique words).

```{r stop-word-lexicons}
count(stop_words)

stop_words %>%
  ggplot(aes(lexicon)) + 
  geom_bar(fill = "darkcyan")

stop_words %>%
  group_by(lexicon) %>%
  count()
  
```

**Q4 (2 marks):** How many *unique* unigrams (distinct words) are there in total across all the cleaned and tokenised transcriptions for each specialty? Which specialty has the largest vocabulary and which the smallest? Complete the code below, then provide a 1-2 sentence interpretation of the output. 

Across the clean and tokenized descriptions for each specialty, Orthopedics had the biggest vocabulary with 7691 unique terms and Radiology had the smallest vocabulary with only 5969 unique terms. Neurology was close behind orthopedics with 7515 distinct words.

```{r q4}
# Uncomment and complete this pipeline-fill in the blank inside n_distinct()
tokenised_data |>
  group_by(medical_specialty) |>
  summarise(unique_words = n_distinct(word))

# un_word <- n_distinct(tokenised_data$word)
# un_id <- n_distinct(tokenised_data$note_id)

# Now calculate the mean number of unique unigrams per NOTE for each specialty
  tokenised_data |>
    group_by(medical_specialty, note_id) |>
    summarise(unique_words = n_distinct(word), .groups = "drop") |>
    group_by(medical_specialty) |>
    summarise(mean_unique_words = mean(unique_words))

```

### Bigrams

We can also combined consecutive words into two-word phrases (bigrams) which better capture medical compound terms that unigrams split apart
(e.g., "blood pressure", "right knee", "general anaesthesia"):

```{r bigrams}
bigrams <- filtered_data |>
  unnest_tokens(bigram, transcription, token = "ngrams",
                n = 2, to_lower = TRUE) |>
  filter(!is.na(bigram)) |>
  mutate(word1 = word(bigram, 1),
         word2 = word(bigram, 2)) |>
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word,
         !str_detect(word1, "\\d"),
         !str_detect(word2, "\\d")) |>
  select(-word1, -word2)

bigrams |>
  count(medical_specialty, bigram, sort = TRUE) |>
  slice_max(n, n = 6, by = medical_specialty, with_ties = FALSE) |>
  arrange(medical_specialty) 
```

**Q5 (1 mark):** What is the most common bigram in each specialty?

The most common bigram in Neurology is "lower extremities". The most common bigram in ORthopedic is "preoperative diagnosis" and the most common bigram in radiology is "carotid artery".

```{r bigram_totals}
bigrams |>
  count(medical_specialty, bigram, sort = TRUE) |>
  slice_max(n, n = 1, by = medical_specialty, with_ties = FALSE) |>
  arrange(medical_specialty) 
```


### Sentence segmentation

Another common text processing step is to split our text into distinct sentences. Again our old friend `unnest_tokens` can do a lot of the heavy lifting!

```{r sentences}
sentences_data <- filtered_data |>
  unnest_tokens(sentence, transcription,
                token = "sentences", to_lower = FALSE)

sentences_data |>
  group_by(medical_specialty) |>
  summarise(total_sentences  = n(),
            n_notes          = n_distinct(note_id),
            mean_per_note    = round(total_sentences / n_notes, 1))
```

**Q6 (1 mark):** How many *unique* sentences are there in each specialty?
Uncomment and complete the code below (replace each `___` with the correct column name):

There are somehow exactly 720 unique sentences in each specialty.

```{r q6}
sentences_data |>
  group_by(medical_specialty) |>
  summarise(total_sentences  = n(),
            unique_sentences = n_distinct(sentences),
            mean_per_note    = round(total_sentences/unique_sentences, 1)
            )

```

## Stemming and Lemmatization

However, given how noisy language is and the many variants of the same word (alternative suffixes) like pluralisation or possessive apostrophes. We also often want to "normalise" words to a common base form. There are two main approaches to this:

- **Stemming**-crude suffix stripping, fast but imprecise:
  "operating" to "operat" (`SnowballC::wordStem`)
- **Lemmatization**-linguistically informed root form:
  "operating" to "operate" (`textstem::lemmatize_words`)
  
  We can apply these normalisations using the associated library functions.

```{r stem-vs-lemma}
word_sample <- tokenised_data |>
  distinct(word) |>
  mutate(stem  = SnowballC::wordStem(word, language = "english"),
         lemma = textstem::lemmatize_words(word))

word_sample |>
  filter(word != stem | word != lemma) |>
  select(word, stem, lemma) |>
  slice_head(n = 15)
```

**Q7 (1 mark):**  Does lemmatising or stemming result in fewer unique words?

Stemming results in fewer unique words. 


```{r stemming_v_lemmatising}
word_sample <- tokenised_data |>
  distinct(word) |>
  mutate(stem  = SnowballC::wordStem(word, language = "english"),
         lemma = textstem::lemmatize_words(word)) 
n_distinct(word_sample$stem)

n_distinct(word_sample$lemma)

```


**Q8 (2 marks):** In 2-3 sentences explain why a general-purpose English
lemmatizer is likely to perform poorly on medical transcription text. Give at
least two examples of medical terms that would likely be handled incorrectly
and describe what goes wrong.

Generally, English is a language with many exceptions, so a general purpose lemmatizer would likely perform poorly on medical texts, especially since many medical terminologies have roots in other languages. Some examples would include villi and hypoxic, which should be lemmatized to villus and hypoxia, but would likely not be changed.

### Applying lemmatisation to the full dataset

Now let's lemmatise the full dataset:

```{r lemmatize, cache=TRUE}
unique_lemmas <- tokenised_data |>
  distinct(word) |>
  mutate(lemma = textstem::lemmatize_words(word))

lemmatised_data <- tokenised_data |>
  left_join(unique_lemmas, by = "word")
```

We can now look more easily at the most frequent (lemmatised) words per specialty
```{r top-lemmas-plot}
lemmatised_data |>
  count(medical_specialty, lemma, sort = TRUE) |>
  slice_max(n, n = 8, by = medical_specialty, with_ties = FALSE) |>
  ggplot(aes(n, reorder_within(lemma, n, medical_specialty),
             fill = medical_specialty)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~medical_specialty, scales = "free_y") +
  scale_y_reordered() +
  labs(title = "Most frequent lemmas per specialty",
       x = "Count", y = NULL)
```

## TF-IDF: Finding Discriminating Terms

Unfortunately, as you can see above, raw counts are dominated by generic medical words ("patient", "history", "left").
TF-IDF (term frequency–inverse document frequency) down-weights terms common
across all documents and up-weights terms characteristic of a specific category.
Sun et al. (2024) used word-frequency analysis as a first step before LDA to
explore the vocabulary of their social work notes.

```{r tfidf}
specialty_tfidf <- lemmatised_data |>
  count(medical_specialty, lemma) |>
  bind_tf_idf(lemma, medical_specialty, n)

specialty_tfidf |>
  group_by(medical_specialty) |>
  slice_max(tf_idf, n = 25, with_ties = FALSE) |>
  ggplot(aes(tf_idf,
             reorder_within(lemma, tf_idf, medical_specialty),
             fill = medical_specialty)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~medical_specialty, scales = "free") +
  scale_y_reordered() +
  labs(title = "Top 25 lemmas by TF-IDF per specialty",
       x = "TF-IDF", y = NULL)
```

**Q9 (2 marks):** Do the top TF-IDF terms make clinical sense for each specialty?
Identify one term that surprises you from any specialty. Complete the code below by replacing
`"your_term_here"` with that term to pull an example transcription containing
it, then briefly explain its clinical context.

The term that surprised me was "spell" in neurology. I expected the clinician to mention something about a patient spelling their name but they used it to describe a patient experiencing certain symptoms (i.e. having a spell).

```{r q9}
filtered_data |>
  filter(str_detect(transcription, "spell")) |>
  slice(1) |>
  pull(transcription) |>
  str_trunc(9000)

```

### Comparing lemma proportions across specialties

Now let's compare across specialties. This plot shows each lemma's proportion in Orthopedic notes (y-axis) against
its proportion in Radiology or Neurology (x-axis, one facet each). Terms near
the dashed diagonal are used similarly in both specialties; terms far from it
are specialty-specific.

```{r proportion-plot, size=12}
specialty_props <- lemmatised_data |>
  count(medical_specialty, lemma) |>
  group_by(medical_specialty) |>
  mutate(proportion = n / sum(n)) |>
  ungroup() |>
  pivot_wider(id_cols     = lemma,
              names_from  = medical_specialty,
              values_from = proportion,
              values_fill = 0) |>
  pivot_longer(c(Radiology, Neurology),
               names_to  = "comparison_specialty",
               values_to = "proportion")

ggplot(specialty_props |> filter(proportion > 0, Orthopedic > 0),
       aes(x = proportion, y = Orthopedic,
           colour = abs(Orthopedic - proportion))) +
  geom_abline(colour = "grey40", linetype = 2) +
  geom_jitter(alpha = 0.1, size = 1.8, width = 0.0003, height = 0.0003) +
  geom_text(aes(label = lemma), check_overlap = TRUE,
            vjust = 1.5, size = 2.5) +
  scale_x_log10(labels = percent_format(accuracy = 0.001)) +
  scale_y_log10(labels = percent_format(accuracy = 0.001)) +
  scale_colour_gradient(limits = c(0, 0.003),
                        low = "darkslategray4", high = "gray75") +
  facet_wrap(~comparison_specialty) +
  theme(legend.position = "none") +
  labs(y = "Orthopedic", x = "Comparison specialty",
       title = "Lemma proportions: Orthopedic vs. Radiology and Neurology")
```

**Q10 (2 marks):** What does this plot tell you about the lexical similarity
between Orthopedic and Neurology compared with Orthopedic and Radiology? Is
this what you would expect given the clinical roles of each specialty? Justify
your answer in 2-4 sentences.

This plot tells me that Orthopedic and Neurology have more terms in common than Orthopedic and Radiology. Based on the clinical roles of each specialty, I would not have anticipated this association. Since orthopedic tends to deal with bones, and many radiology charts would have to do with x-rays for broken or potentially broken bones. Orthopedic and Neurology likely have so many similar words because Neurology has such a broad scope that many of the words overlap with orthopedics, which also has a larger scope than Radiology.

---

## Topic Modelling with Latent Dirichlet Allocation

Latent Dirichlet Allocation (LDA) is an unsupervised probabilistic model that
treats each document as a mixture of topics and each topic as a probability
distribution over words. Unlike supervised classifiers, LDA discovers thematic
structure from the text alone-no specialty labels are used during training.
This mirrors the approach of Sun et al. (2024), who applied LDA to clinical
social work notes to uncover latent SDoH themes without predefined categories.

Two key outputs of a fitted LDA model:

| Output | Symbol | Interpretation |
|---|---|---|
| Word–topic matrix | β (beta) | Probability that word *w* is generated by topic *k* |
| Document–topic matrix | γ (gamma) | Proportion of document i.e., transcript *d* attributed to topic *k* |

### Building the document-term matrix

We aggregate lemma counts per note and apply two filters before building the DTM:

1. **Minimum document frequency** - drop terms appearing in fewer than 5 notes to reduce sparsity.
2. **TF-IDF filtering** - drop terms whose mean TF-IDF (computed per note) falls in the bottom quartile. Words that are common across nearly all notes have low IDF and therefore low TF-IDF; retaining them would give LDA noisy, non-discriminative signal.

```{r dtm-build, cache=TRUE}
term_counts <- lemmatised_data |>
  count(note_id, lemma)

# Keep only terms present in >= 5 notes
doc_freq_filter <- term_counts |>
  distinct(note_id, lemma) |>
  count(lemma, name = "doc_count") |>
  filter(doc_count >= 5)

# Keep only terms whose mean per-note TF-IDF is above the 25th percentile
tfidf_filter <- term_counts |>
  bind_tf_idf(lemma, note_id, n) |>
  group_by(lemma) |>
  summarise(mean_tfidf = mean(tf_idf), .groups = "drop") |>
  filter(mean_tfidf > quantile(mean_tfidf, 0.25))

emr_dtm <- term_counts |>
  semi_join(doc_freq_filter, by = "lemma") |>
  semi_join(tfidf_filter,    by = "lemma") |>
  cast_dtm(note_id, lemma, n)

emr_dtm <- emr_dtm[rowSums(as.matrix(emr_dtm)) > 0, ]

cat("DTM:", nrow(emr_dtm), "documents ×", ncol(emr_dtm), "terms\n")
```

### Fitting the LDA model

We fit LDA with `k = 3` topics-one per specialty-so we can directly compare
the discovered topics against the known specialty labels.  Again this would likely be something you explore or optimise using more sophisticated hierarchical models.

```{r fit-lda, cache=TRUE}
set.seed(42)
k_topics <- 3
emr_lda  <- topicmodels::LDA(emr_dtm, k = k_topics,
                              control = list(seed = 42))
```

### Word–topic probabilities (β)

The β matrix tells us which words are most characteristic of each topic:

```{r lda-beta-plot}
lda_beta <- tidy(emr_lda, matrix = "beta")

lda_beta |>
  group_by(topic) |>
  slice_max(beta, n = 12, with_ties = FALSE) |>
  ggplot(aes(beta,
             reorder_within(term, beta, topic),
             fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free_y",
             labeller = labeller(topic = function(x) paste("Topic", x))) +
  scale_y_reordered() +
  labs(title = "Top 12 words per LDA topic (β probabilities)",
       x = "β  (word–topic probability)", y = NULL)
```

**Q11 (2 marks):** Based on the top words in the β plot, give each topic a brief
descriptive clinical label (e.g., "imaging", "musculoskeletal", "neurological").
Which topic do you think corresponds most closely to which specialty, and why?
Does the vocabulary in each topic match your expectations from the TF-IDF
results in Section 6?

Topic 1: Imaging - Radiology because all the terms are related to things that are frequently seen in medical imaging, like "ct" or "mass". 
Topic 2: Spines and vertebral columns - Neurology because the spinal cord is a very important nerve.
Topic 3: Orthopedic injuries - Orthopedic because these are all terms used for orthopedic interventions and refer to muscles and tendons. 

### Document–topic distributions (γ) by specialty

The γ matrix tells us how much each document (i.e., transcript in this case) draws from each topic. We join it
with specialty labels to see whether LDA-discovered topics align with the known
clinical categories:

```{r lda-gamma}
lda_gamma <- tidy(emr_lda, matrix = "gamma")

lda_gamma_specialty <- lda_gamma |>
  left_join(
    filtered_data |>
      mutate(document = as.character(note_id)) |>
      select(document, medical_specialty),
    by = "document"
  )
```

```{r lda-gamma-plot}
lda_gamma_specialty |>
  group_by(medical_specialty, topic) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop") |>
  mutate(topic = factor(topic, labels = paste("Topic", seq_len(k_topics)))) |>
  ggplot(aes(topic, mean_gamma, fill = medical_specialty)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Mean topic proportion by specialty (γ)",
       x = NULL, y = "Mean γ", fill = "Specialty")
```

**Q12 (3 marks):**

(a) Based on the γ plot, which specialty has the clearest dominant topic (i.e.,
one topic has a noticeably higher mean γ than the others for that specialty)?
Does this align with the clinical label you assigned to that topic in Q11?

Topic 3 is the most clearly associated with orthopedics, this aligns with the label I assigned in Q11.

(b) Uncomment and complete the code below to compute, for each specialty, the single topic
with the highest mean γ. Replace each `___` with the correct column name or
expression.

```{r q12b}
lda_gamma_specialty |>
 group_by(medical_specialty, topic) |>
 summarise(mean_gamma = mean(gamma), .groups = "drop") |>
 group_by(topic) |>
 slice_max(mean_gamma, n = 1)
```

(c) Sun et al. (2024) ran LDA on social work notes *without knowing in advance*
what topics would emerge. How does their situation differ from this exercise,
where we set `k = 3` to match three known specialties? What does this suggest
about how to choose the number of topics *k* in a real exploratory analysis? \n

This situation is different because we know how many groups we are expecting. Sun et al. did not know how many topics or groups would emerge from their analysis. This suggests that we choose *k* based on the approximate number of topics that emerge.

---

## LLMs for Clinical Text Analysis

Finally, let's address the LLM in the room.
Chung et al. (2024) showed that GPT-4 Turbo, with no task-specific training,
could classify perioperative outcomes from pre-operative notes using carefully
designed prompts (zero-shot, few-shot, chain-of-thought). 

Let's experiment similarly to classify a single note and then compare
its reasoning with the LDA topic model's assignment. You can use whichever LLM you are comfortable using (Dalhousie students
have access to Microsoft Copilot via Office 365 (copilot.microsoft.com) with some degree of data protection/agreements).

### Sample note

```{r sample-note, cache=FALSE}
# Sample from notes that appear in the topic model
set.seed(42)
dtm_note_ids <- as.integer(rownames(emr_dtm))

q10_note <- filtered_data |>
  filter(note_id %in% dtm_note_ids) |>
  slice_sample(n = 1)

cat("=== True specialty:", q10_note$medical_specialty, "===\n\n")
cat("Description:", q10_note$description, "\n\n")
cat("--- Transcription (first 1200 characters) ---\n")
cat(str_trunc(q10_note$transcription, 1200))
```

LDA topic assignment for this note (γ values, highest first):

```{r sample-note-gamma, cache=FALSE}
lda_gamma_specialty |>
  filter(document == as.character(q10_note$note_id)) |>
  mutate(topic = paste("Topic", topic)) |>
  arrange(desc(gamma))
```

**Q13 (3 marks):** Copy the transcription above into Microsoft Copilot and use
the following prompt:

> *"I will give you a medical transcription. Classify it as exactly one of:
> Orthopedic, Radiology, or Neurology. Think step by step and explain your
> reasoning before giving your final answer. Transcription: [paste text here]"*

(a) What specialty did the LLM predict, and was it correct? Does it match the
topic with the highest γ from the LDA model above?

The LLM predicted the correct specialty (orthopedic). Yes, it matches topic 3 which has the highest gamma from the LDA model.

(b) What specific features of the text did the LLM cite to justify its
classification?

The LLM cited medical terms that related to conditions and treatments typical of orthopedics (i.e. Tailor's bunionette and bunionectomy)

(c) Compare LLM's explanation-based classification with LDA's unsupervised
topic assignment. Which approach would you prefer for classifying a large batch
of new clinical notes, and why? 

I would prefer the LDA model for a large batch of clinical notes because it would still likely classify the notes correctly and would use significantly less computing resources and power and would likely be quicker.

---

## Tying back to the literature

**Q14 (1 mark):** Sun et al. (2024) applied LDA topic modelling
to clinical social work notes to explore SDoH factors. Name two of the
SDoH-related topics they identified and state the name of the metric they used
to evaluate topic quality (hint: it measures the coherence of the top words
within a topic).

Two topics they identified were Mental Health and Family. They used Topic Coherence to measure the degree of  similarity between words in a topic.

**Q15 (1 mark):** Chung et al. (2024) tested GPT-4 Turbo on both
classification and numerical prediction tasks. Name one task where the LLM
performed well (give the F1 score) and one where it failed to beat a simple
baseline (give the MAE and the baseline value for comparison).

The LLM did a good job in prediction tasks, for example hospital admissions. The F1 score for hospitalization improved to 0.4 from 0.3. It did not do well in making numerical predictions. An example of this would be hospital duration with an MAE of 4.55 days and a baseline MAE of 5.4 days.

---

## Reproducibility

```{r session-info, cache=FALSE}
sessioninfo::session_info()
```

---

*Dataset: MTSamples via `clinspacy` (mtsamples.com) with material drawn from
Silge & Robinson,* Text Mining with R *(tidytextmining.com)*
