---
title: "Practical 3 - Medical Imaging: Detecting Diabetic Retinopathy with CNNs"
author: "Sarah de Lima"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
    theme: cosmo
    highlight: tango
    number_sections: false
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110,
  cache      = TRUE,
  cache.lazy = FALSE
)

GLOBAL_SEED <- 42
set.seed(GLOBAL_SEED)
```

> **Submission.** Knit your Rmd file into HTML and submit it via Brightspace. Include a link to your git repository (GitHub or git.cs.dal.ca) as a comment on your submission.

---

## 0 Setup

Unlike the tabular and text modules, deep learning on images leans on libraries written
in C++/CUDA. These can be a bit challenging to install and use via R but the [`mlverse`](https://mlverse.org/)
stack is a modern and supposedly **self-contained and cross-platform** set of R bindings for some of the main deep learning/image libraries:

- `torch` is one of the main current neural network library (the other being `tensorflow`) and provides lots of functions for defining neural-network layers and converting them to efficient code for training and running on GPUs/CPUs.  The R library is really just a set of R functions which let you run underlying C++/CUDA from within R (these are known as `bindings`). When you install `torch` in R, it downloads its own copy of LibTorch on first use.
- `torchvision` is a set of efficient helper functions for handling images in neural networks (e.g., image transforms, dataset helpers, and access to *pre-trained* networks).
- `magick` is a tool for viewing and manipulating images which works on Mac, Windows, or Linux. These tasks can be surprisingly operating system specific usually.

Copy the following into the console to install dependencies (make sure it finishes correctly):
```
install.packages(c(
  "torch", "torchvision", "luz",   # deep learning stack
  "tidyverse", "here", "fs",       # data wrangling + reproducible paths
  "magick", "gridExtra",           # image display
  "yardstick"                      # evaluation metrics (also used in Lab 1)
))
```
Then copy the following into the console to finish the `torch` install (it will take a few minutes to complete):
```
install_torch()
```

If that all worked, the following should run without an error as a normal code-block:

```{r packages}
library(torch)        # tensors, autograd, neural-network layers
library(torchvision)  # image transforms, datasets, pretrained models
library(luz)          # high-level training loop (fit/predict/metrics)
library(tidyverse)    # dplyr, ggplot2, readr, purrr, ...
library(here)         # reproducible file paths
library(fs)           # filesystem helpers
library(magick)       # image display only
library(gridExtra)    # arranging plots in a grid
library(yardstick)    # ROC-AUC, sensitivity, specificity, confusion matrix

torch_manual_seed(GLOBAL_SEED)
set.seed(GLOBAL_SEED)
```

## 1 Introduction

Medical imaging is one of the most data-rich corners of medicine with a single trip to the emergency room able to
 generate hundreds of gigabytes across X-ray, CT, MRI, PET and ultrasound. As covered in
the lecture, the *shape* of that data ranges from a single number (e.g., optical density) up to 4D
tensors (a PET-CT time-series), and the *questions* we ask ranges from preprocessing and
segmentation through to detection, classification and outcome **prediction**.

This practical focuses on **2D colour fundus photographs (CFPs)** of the retina and the task
of detecting **diabetic retinopathy (DR)** i.e., damage to the retinal blood vessels caused by
diabetes.  This is one of the leading global causes of preventable blindness. 
Manual screening is done accurately by ophthalmologists but does not easily scale (especially across resource-limited settings) to the ~500 million people living with diabetes. Predictive models may help fill this scaling gap.

We will very loosely mirror the analyses in the two assigned papers and refer back to them in the questions:

- **Gulshan et al. (2016, *JAMA*)** trained an Inception-v3 CNN (with ImageNet transfer
  learning) to **detect** *referable* DR, reaching an AUC of ~0.99 and reporting two clinical
  operating points (one tuned for high sensitivity, one for high specificity).
- **Arcadu et al. (2019, *npj Digital Medicine*)** went further and tried to **predict future**
  DR progression from a single visit, using field-specific Inception-v3 networks aggregated by
  a random forest, and interpreted them with attribution maps and SHAP.

**Learning objectives:** by the end of this practical you should be able to:

- Load, inspect and normalise medical image data as `torch` tensors
- Build, train and evaluate a convolutional neural network (CNN) from scratch with `luz`
- Apply data augmentation and explain its role as a prior on image variability
- Use **transfer learning** to apply an ImageNet-pretrained network to the problem
- Evaluate a clinical classifier with ROC-AUC, sensitivity/specificity and **operating points**
- Connect model performance to the real-world hurdles of clinical deployment

## 2 The Data

### 2.1 Download and cache

We use a curated subset of fundus photographs graded for DR severity on the 5-point
International Clinical Diabetic Retinopathy (ICDR) scale (the clinical cousin of the ETDRS
scale used by Arcadu et al.):

| Grade | Severity |
|---|---|
| 0 | No DR |
| 1 | Mild non-proliferative DR |
| 2 | Moderate non-proliferative DR |
| 3 | Severe non-proliferative DR |
| 4 | Proliferative DR |

As in Lab 1, we download a prepared archive once and cache it locally so the chunk only
re-runs if the file is missing.

```{r data-paths}
data_dir  <- here()
cache_dir <- fs::path(data_dir, "cache")
fs::dir_create(cache_dir)

data_url <- "https://maguire-lab.github.io/health_data_science_research_2026/static_files/practicals/dr_fundus.zip"
zip_path <- fs::path(cache_dir, "dr_fundus.zip")
img_root <- fs::path(data_dir, "dr_fundus")
```

```{r download-data, cache = TRUE}
if (!fs::dir_exists(img_root)) {
  if (!fs::file_exists(zip_path)) {
    download.file(data_url, zip_path, mode = "wb", quiet = TRUE)
  }
  unzip(zip_path, exdir = data_dir)
}

# The archive contains:
#   dr_fundus/images/<id>.jpg   -- the fundus photographs
#   dr_fundus/labels.csv        -- columns: image, split, grade
fs::dir_ls(img_root)
```

### 2.2 Load the labels

```{r load-labels}
labels <- read_csv(fs::path(img_root, "labels.csv")) |>
  mutate(
    path  = fs::path(img_root, "images", image),
    grade = as.integer(grade)
  ) |>
  # Simplify to the two extremes of the ICDR scale: grade 0 vs grade 4
  filter(grade %in% c(0L, 4L)) |>
  mutate(
    grade     = factor(grade, levels = c(0, 4)),
    # binary target: 1 = severe (grade 4, certainly referable), 0 = healthy (grade 0)
    referable = as.integer(grade == 4)
  )

glimpse(labels)
```

To keep this teaching example clean and the dataset as small as possible, we frame the task as the **easiest possible** version of
the clinical problem: telling the two *extremes* apart - a **clearly-healthy** eye (grade 0, no
DR) versus the **most severe** (grade 4, proliferative DR).   We will NOT use cross-validation this time and instead use a separate validation and test dataset.  This reduces the number of times we have to train these models.

**Q1 (2 marks):** In your own words, what is diabetic retinopathy, and what is the clinical benefit of being able to flag *referable* DR automatically from a photograph? Why might a screening tool prefer to err towards over-referral? *(Hint: see the opening paragraphs of Gulshan et al.)*

*Diabetic retinopathy is an eye condition associated with diabetes. It's when  increased blood sugar damages the capillaries in they eye, causing bleeding or lesions. It would be especially important to flag referable DR because it is more severe and can threaten an individual's sight. Screening tools can over-refer because Machine Learning is often trained to detect any lesion or abnormality, and label any level of DR as referable DR*.

**Q2 (2 marks):** Based on `glimpse()` and the data dictionary above, what information about
how these images were captured is **missing** that could matter for an analysis? Give two examples and say why each matters. \n
*Information about the technology used to capture the images is missing, this would be important because different equipment may have different light and capture settings, which could impact how the algorithm picks up on edges and abnormalities. There is also information missing about patient's medical history, for example if they have cataracts. The presence of cataracts would impact the quality of the retinal image, and could impact how the algorithm identifies abnormalities.*


### 2.3 Class balance

```{r class-balance, fig.height = 3}
labels |>
  mutate(label = if_else(referable == 1, "Severe DR", "Healthy")) |>
  count(split, label) |>
  ggplot(aes(label, n, fill = label)) +
  geom_col() +
  facet_wrap(~split) +
  scale_fill_manual(values = c("Healthy" = "grey60",
                               "Severe DR"  = "firebrick")) +
  labs(x = NULL, y = "Number of Images",
       title = "Healthy vs Severe DR Balance per Split") +
  theme(legend.position = "none")
```

**Q3 (1 marks):** What sort of ML issues may arise with a dataset that has this sort of label/class balance?

*Since the data is imbalanced, it can overweight the overrepresented sample, so it may misclassify samples as healthy.*

### 2.4 Look at the images

It is *always* worth looking directly at the raw data before modelling. Each time you re-run this block you will get a different random sample of images.

```{r show-images, fig.width = 8, fig.height = 4.5}
samples <- labels |>
  filter(split == "train") |>
  slice_sample(n = 6)

plots <- map2(samples$path, samples$grade, \(p, g) {
  magick::image_ggplot(magick::image_read(p)) +
    ggtitle(paste("Grade", g)) +
    theme(plot.title = element_text(size = 9))
})

gridExtra::grid.arrange(grobs = plots, nrow = 2, ncol = 3)
```

**Q4 (3 marks):** Looking at these images, identify **two** properties that vary between
photographs of the same class that we may want to normalise/process before training?  Can you see any consistent differences between Grade 4 and Grade 0 photographs? (Rerun the code-block a few times to get different images).

*I would want to adjust the lighting as there are very dark images where it is very difficult to distinguish any lesions. I would also adjust the warmth of the images (how red-tinted an image is) because that might affect how light is perceived by the model and thus the outcome.* 
---

## 3 Pre-processing

### 3.1 From image file to tensor

A colour image is a **3D tensor** of shape `channels × height × width`. `torch` expects a 4D batch of shape
`images × channels × height × width`. We build a small, *lazy* pipeline that reads each file only when needed
(so we never hold the whole dataset in memory at once - although this specific dataset is small enough to do that on most computers) and:

1. Loads the JPEG into a `[0, 1]` tensor,
2. Resizes it to 224 × 224 (most neural networks require a single consistent input size),
3. Normalises each channel using the mean colour intensity and standard deviation across the entire **ImageNet** dataset.

We use `ImageNet` numbers in step 3 because in Section 6 we reuse a network pretrained on ImageNet, and a
pretrained network expects its inputs scaled the same way as the training data.

```{r transforms}
imagenet_mean <- c(0.485, 0.456, 0.406) # manually pulled from the internet... "magic numbers" that you probably shouldn't trust normally
imagenet_std  <- c(0.229, 0.224, 0.225)

# transform for evaluation / training-from-scratch: deterministic resize + normalise
eval_transform <- function(path) {
  base_loader(path) |>            # H x W x C array in [0, 1]
    transform_to_tensor() |>      # -> C x H x W float tensor
    transform_resize(c(224, 224)) |>
    transform_normalize(mean = imagenet_mean, std = imagenet_std)
}
```

### 3.2 A `torch` dataset and dataloaders

We need to do some special code to load our transformed data into a `torch` network for training.
For efficiency, we train neural networks using "batches" of training data. This means we have it make predictions on a set of images/inputs all together and then update all the weights at once based on how well it did across the batch as a whole.

```{r dataset}
fundus_dataset <- dataset(
  name = "FundusDataset",
  initialize = function(df, transform) {
    self$paths     <- df$path
    # torch uses 1-based class indices, so referable {0,1} -> {1,2}
    self$y         <- torch_tensor(df$referable + 1L, dtype = torch_long())
    self$transform <- transform
  },
  .getitem = function(i) {
    list(x = self$transform(self$paths[i]), y = self$y[i])
  },
  .length = function() {
    length(self$paths)
  }
)

train_df <- labels |> filter(split == "train")
valid_df <- labels |> filter(split == "valid")
test_df  <- labels |> filter(split == "test")

train_ds <- fundus_dataset(train_df, eval_transform)
valid_ds <- fundus_dataset(valid_df, eval_transform)
test_ds  <- fundus_dataset(test_df,  eval_transform)

batch_size   <- 16
train_dl <- dataloader(train_ds, batch_size = batch_size, shuffle = TRUE)
valid_dl <- dataloader(valid_ds, batch_size = batch_size, shuffle = FALSE)
test_dl  <- dataloader(test_ds,  batch_size = batch_size, shuffle = FALSE)
```

Let's confirm the shape of one batch:

```{r batch-shape}
one_batch <- train_dl |> dataloader_make_iter() |> dataloader_next()
dim(one_batch$x)   # expect: batch_size x 3 x 224 x 224
```

**Q5 (2 marks):** Explain what each of the four numbers in the batch dimension means. Why is
the channel dimension `3` here? If, instead of CFP images, we were using a 3D MRI grayscale image that was 1000 pixels/voxels in each dimension, what would be the dimensions output by the above command?

*Batch size is the number of images in the set, the 3 represents the number of channels (this is a 3 channel colour image) and the 224x224 would be the height and width of each image in pixels. If we were using the measurements outlined in the question, the dimensions would be 16 x 1 x 1000 X 1000*

**Q6 (1 marks):** If we didn't want to use pre-trained ImageNet networks then we would still probably want to normalise the intensities in our dataset.  What issue might we cause if we normalised using the mean and standard deviation of the pixel intensities across the *whole* dataset (i.e., train, test, and validate combined)?

*We risk data leakage, the result of which is an overfitted model with falsely high performance statistics that is not generalizable to other datasets.*

---

## 4 A CNN from Scratch

We first train a small CNN *from scratch*, with no outside knowledge. 
Convolutions slide learnable filters across the image to capture local spatial
patterns (see lecture); stacking convolutional layers, non-linearities (sigmoid/ReLUs), and pooling blocks builds up "increasingly higher-order learnt representations." 

### Why convolutions?

A fully-connected layer would give every input pixel its own weight to every output unit. For a
224×224×3 image that is ~150,000 inputs *per output unit* far too many parameters to learn
from a few thousand images.  This would also ignore the important concept that pixels which are **near each
other** in an image are related to one another. A **convolutional layer** enforce two structural assumptions that align with how
images actually work:

- **Locality.** Each output looks at only a small patch of the input - here a `kernel_size = 3`
  filter sees a 3×3 neighbourhood at a time. A learnable filter is just a small grid of weights
  that is multiplied element-wise with the patch and summed to a single number; sliding it
  across the whole image produces a *feature map* highlighting wherever the pattern encoded in the convolutional filter (an edge,
  a blob, a particular colour gradient) occurs.
- **Weight sharing.** The *same* filter is reused at every position. So a filter that has
  learned to detect, say, a microaneurysm-like dark spot will fire wherever that spot appears -
  the network does not have to relearn it for every corner of the image. The fancy term for this is **translation equivariance**.

Each conv layer learns *many* filters at once — the second number in `nn_conv2d(3, 16, …)` is
the number of output filters, so this layer turns the 3 colour channels into **16 feature
maps**. As we go deeper we ask for more filters (16 to 32 to 64) because later layers compose the
simple early patterns into rarer, higher-order ones (e.g., edges to textures to specific lesion-shaped motifs).
Each conv is followed by a `nn_relu()`, a simple `max(0, x)` nonlinearity like any normal neural network.

### Why pooling?

`nn_max_pool2d(2)` slides a 2×2 window across each feature map and keeps only the **maximum**
value in each window, halving the height and width (so 224 to 112 to 56 …). This does three useful
things:

- **Translation *invariance*.** Once a filter has reported "an edge is present somewhere in this
  patch," we often do not care about its exact pixel position. Keeping the strongest response and
  discarding precise location makes the network robust to the lesion being shifted by a few
  pixels.  This is useful, since fundus images are rarely perfectly aligned to one another (i.e., "registered").
- **A growing receptive field.** After pooling, each value summarises a larger region of the
  original image, so the *next* 3×3 filter effectively "sees" a bigger area. This is how a stack
  of small filters can end up responding to large structures.
- **Cheaper computation.** Halving the resolution quarters the number of values the next layer
  must process, which is what lets us afford more filters as we go deeper.

The final `nn_adaptive_avg_pool2d(c(1, 1))` is **global average pooling**: it collapses each of
the 64 feature maps to a single number corresponding to its average activation over the whole image. This
length-64 summary vector finally feeds into the classifier part of network.

We define this "simple" CNN using the following code:

```{r scratch-cnn}
simple_cnn <- nn_module(
  "SimpleCNN",
  initialize = function() {
    self$features <- nn_sequential(
      nn_conv2d(3, 16, kernel_size = 3, padding = 1), nn_relu(), nn_max_pool2d(2),
      nn_conv2d(16, 32, kernel_size = 3, padding = 1), nn_relu(), nn_max_pool2d(2),
      nn_conv2d(32, 64, kernel_size = 3, padding = 1), nn_relu(),
      nn_adaptive_avg_pool2d(c(1, 1))   # global average pool -> 64 x 1 x 1
    )
    self$classifier <- nn_sequential(
      nn_dropout(p = 0.5),
      nn_linear(64, 2)                  # 2 logistic functions to get the probability of grade 0 and grade 4
    )
  },
  forward = function(x) {
    x <- self$features(x)
    x <- torch_flatten(x, start_dim = 2)
    self$classifier(x)
  }
)
```

We then use `luz` to wrap the training loop. It minimise cross-entropy and tracks accuracy on the separate
validation set each time we train on all of our training data batches (also known as an epoch).

```{r train-scratch, cache = TRUE}
fitted_scratch <- simple_cnn |>
  setup(
    loss      = nn_cross_entropy_loss(),
    optimizer = optim_adam,
    metrics   = list(luz_metric_accuracy())
  ) |>
  fit(train_dl, epochs = 5, valid_data = valid_dl, verbose = FALSE)
```

```{r scratch-curves, fig.height = 3.2}
luz::get_metrics(fitted_scratch) |>
  ggplot(aes(epoch, value, colour = set)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Training-from-scratch learning curves",
       x = "Epoch", y = NULL, colour = NULL)
```

**Q7 (2 marks):** Look at the validation accuracy and loss across epochs. Do you think small network is actually
*learning* to separate referable from non-referable eyes, why or why not? 
*No, I don't think the neural network is actually learning to separate referable from non-referable eyes because even though it is accurate on the training dataset, it does not improve after each epoch for the validation set. When looking at the loss of the model, it does improve with the validation set, but not in the same way that it does with the training set, which indicates that the model is not performing the same in both groups.*


## 5 Data Augmentation

When real training data is scarce, **augmentation** synthesises plausible new examples by
applying random transformations (i.e., inducing a prior on how we expect images to vary).

```{r augmentation-demo, fig.width = 8, fig.height = 3}
demo_path <- train_df$path[1]
orig <- base_loader(demo_path) |> transform_to_tensor()

show_tensor <- function(t, title) {
  arr <- as.array(t$permute(c(2, 3, 1)))           # C,H,W -> H,W,C for plotting
  arr <- (arr - min(arr)) / (max(arr) - min(arr))  # rescale to [0,1] for display
  magick::image_ggplot(magick::image_read(as.raster(arr))) +
    ggtitle(title) + theme(plot.title = element_text(size = 9))
}

aug1 <- orig |> transform_random_horizontal_flip(p = 1)
aug2 <- orig |> transform_color_jitter(brightness = 0.4, contrast = 0.4)

gridExtra::grid.arrange(
  show_tensor(orig, "Original"),
  show_tensor(aug1, "Horizontal flip"),
  show_tensor(aug2, "Colour jitter"),
  nrow = 1
)
```

We add augmentation **only to the training pipeline** (never to validation/test - we evaluate
on untouched images to avoid overfitting):

```{r train-transform}
train_transform <- function(path) {
  base_loader(path) |>
    transform_to_tensor() |>
    transform_random_resized_crop(size = c(224, 224), scale = c(0.8, 1.0)) |>
    transform_random_horizontal_flip(p = 0.5) |>
    transform_color_jitter(brightness = 0.2, contrast = 0.2) |>
    transform_normalize(mean = imagenet_mean, std = imagenet_std)
}
```

**Q8a (2 marks):** These augmentations have some real potential drawbacks especially with data from eyes! Explain why flipping an eye horizontally may be misleading in this context then copy the above code to create a new augmentation pipeline `train_transform_fixed` which removes the horizontal flip and adds
a *small random rotation* (`transform_random_rotation(degrees = ...)`). (Hint: use `?transform_random_rotation` to figure out the parameters for this function).

*Flipping an eye may be misleading because most humans have a left and right eye, and flipping one image could over represent either left or right.*

```{r q8-augment}
# Uncomment, complete the rotation, and rebuild the augmented training dataloader.
train_transform <- function(path) {
  base_loader(path) |>
    transform_to_tensor() |>
    transform_random_resized_crop(size = c(224, 224), scale = c(0.8, 1.0)) |>
    #transform_random_horizontal_flip(p = 0.5) |>
    transform_random_rotation(degrees = c(-45, 45)) |>
    transform_color_jitter(brightness = 0.2, contrast = 0.2) |>
    transform_normalize(mean = imagenet_mean, std = imagenet_std)
}
```

Now apply your new augmentations - this will use the incorrect augmentations if you haven't completed Q8!

```{r apply_augments}
train_ds_aug <- fundus_dataset(train_df, train_transform)
train_dl_aug <- dataloader(train_ds_aug, batch_size = batch_size, shuffle = TRUE)
```

Now we re-train the *exact same* `simple_cnn` from Section 4, changing only the data it
sees: the augmented `train_dl_aug` in place of the un-augmented `train_dl`. Everything else
(architecture, loss, optimiser, epochs) is held fixed so the comparison is like-for-like.

```{r train-scratch-aug, cache = TRUE}
fitted_scratch_aug <- simple_cnn |>
  setup(
    loss      = nn_cross_entropy_loss(),
    optimizer = optim_adam,
    metrics   = list(luz_metric_accuracy())
  ) |>
  fit(train_dl_aug, epochs = 5, valid_data = valid_dl, verbose = FALSE)
```

```{r scratch-aug-curves, fig.height = 3.2}
bind_rows(
  luz::get_metrics(fitted_scratch)     |> mutate(run = "No Augmentation"),
  luz::get_metrics(fitted_scratch_aug) |> mutate(run = "Augmentation")
) |>
  filter(set == "valid") |>
  ggplot(aes(epoch, value, colour = run)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Simple CNN: validation curves with vs without augmentation",
       x = "Epoch", y = NULL, colour = NULL)
```

**Q8b (2 marks):** Compare the validation curves of the augmented and un-augmented runs of the
*same* network. Did augmentation help, hurt, or make little difference here, and why might that
be the case for a dataset and network of this size? \n

*Augmentation made no difference in the accuracy of the model and made little difference in the loss of the model. This would be the case because we have a very small sample size, and augmentation would not  increase the sample size enough for the model to improve sufficiently. *

## 6 Transfer Learning

Training from scratch on a few hundred images is hopeless. The breakthrough in **both**
assigned papers was **transfer learning**: start from a network already trained on the
1.2-million-image, 1000-class [ImageNet](https://www.image-net.org/) dataset, keep its learned
edge/texture/shape detectors, and only re-train the classification part of the network for our task. 

Here we load a pretrained **ResNet-18**, freeze its convolutional backbone, and replace its
final fully-connected layer with a fresh 2-class head.

```{r transfer-model}
train_ds_aug <- fundus_dataset(train_df, train_transform)
train_dl_aug <- dataloader(train_ds_aug, batch_size = batch_size, shuffle = TRUE)


resnet_transfer <- nn_module(
  "ResNetTransfer",
  initialize = function() {
    self$model <- model_resnet18(pretrained = TRUE)
    # freeze the backbone: only the new head will be trained
    for (par in self$model$parameters) par$requires_grad_(FALSE)
    n_in <- self$model$fc$in_features
    self$model$fc <- nn_linear(n_in, 2)   # new, trainable classification head
  },
  forward = function(x) {
    self$model(x)
  }
)
```

```{r train-transfer, cache = TRUE}
fitted_tl <- resnet_transfer |>
  setup(
    loss      = nn_cross_entropy_loss(),
    optimizer = optim_adam,
    metrics   = list(luz_metric_accuracy())
  ) |>
  fit(train_dl_aug, epochs = 5, valid_data = valid_dl, verbose = FALSE)
```

```{r transfer-curves, fig.height = 3.2}
luz::get_metrics(fitted_tl) |>
  ggplot(aes(epoch, value, colour = set)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Transfer-learning learning curves",
       x = "Epoch", y = NULL, colour = NULL)
```

**Q9 (2 marks):** What is ImageNet, and why would convolution and pooling layers learnt on photos of cats, cars
and mushrooms improve our ability to predict DR from CFP images (none of which are in the ImageNet dataset)?

*ImageNet is a database of images that is often used in training visual recognition software. Convolution and pooling layers learnt on the generic photos on ImageNet would be useful in predicting DR because the model would still learn how to detect edges and spatial relations. Pooling condenses the feature map and allows for the model to reduce dimensionality.*

## 7 Evaluating Like a Clinician

Accuracy is misleading under class imbalance (Lab 1... may help with an earlier question in this lab...). Clinically, and in both papers, models are judged on the **ROC curve**, the **AUC**, and a choice of **operating point** trading off
sensitivity (catching disease) against specificity (avoiding false alarms).

We get predicted probabilities of grade 4 DR on the held-out test set:

```{r predict-test}
test_logits <- predict(fitted_tl, test_dl)
test_probs  <- as.numeric(as.array(nnf_softmax(test_logits, dim = 2)[, 2]))

eval_df <- test_df |>
  mutate(
    .pred_referable = test_probs,
    truth = factor(referable, levels = c(1, 0),
                   labels = c("Referable", "Not referable"))
  )
```

```{r roc, fig.width = 5, fig.height = 4}
yardstick::roc_auc(eval_df, truth, .pred_referable, event_level = "first")

yardstick::roc_curve(eval_df, truth, .pred_referable, event_level = "first") |>
  autoplot() +
  labs(title = "ROC curve: referable DR (test set)")
```

A model with two outputs gives us a whole curve, so we can pick *operating points*. Gulshan
et al. reported one tuned for **high specificity** and one for **high sensitivity**.

**Q10 (2 marks):** Define sensitivity and specificity and explain how there is usual a trade-off between them.  Which is more important in clinical screening program? Approximated using this ROC curve (dark line in plot above - just eyeball the value using the plot), how much specificity would you lose if you wanted perfect sensitivity? Similarly, how bad would the sensitivity be with perfect specificity? 

*Sensitivity is the proportion of positive cases that are correctly identified as positive. Specificity is the proportion of negative cases that are correctly identified as negative cases. Which one would be more important would depend entirely on what was being screened for. For DR, we would likely want a more specific test because misclassifying it as a false negative would likely have severe consequences for the patient. If you wanted perfect sensitivity, the specificity would likely be about 40% and if you wanted perfect specificity, the sensitivity would likely only be about 30%.*

## 8 Interpretability and Clinical Translation

Good performance metrics are necessary but nowhere near sufficient for clinical use.

**Interpretability.** Arcadu et al. did not stop at a number — they produced **attribution
maps** (via guided back-propagation) showing *where* each network looked, and used **SHAP** to
see which retinal fields drove the random-forest aggregation. Reassuringly, the maps lit up on
microaneurysms, haemorrhages and exudates - the same lesions clinicians use.  
Unfortunately, these sort of analysis methods proved a little too challenging to get working in a robust cross-platform way for this lab!

**Q11 (1 marks):** Why is an attribution/saliency map more convincing evidence than test
accuracy/sensitivity/specificity alone when arguing a model is clinically trustworthy? Give one way a model could
achieve high performance metrics for the *wrong* reason that an attribution map might expose

*Attribution maps can be more convincing than test metrics alone because they can identify which decision mechanisms a model is using, which can be compared to how humans might make that decision. A model may achieve high performance metrics if it is measuring a confounding variable. For example if all examples of DR happened to be in the right eye for a dataset, the model may be relying on positioning of the fovea and optical nerve to diagnose DR rather than the presence of lesions.*

**Q12 (1 marks):** Gulshan et al. validated on **two** datasets, including the publicly
available Messidor-2 (a *different* country, cameras and protocol). Why is this kind of
*external* validation so much stronger evidence than a held-out split of the same data? 

*External validation is stronger evidence than a held-out split of the same data because the characteristics of an external dataset will be slightly different even if the data is similar. When models are trained on one dataset, they can learn how to perform on that data, but might perform poorly on data with slightly different attributes or without any potentially unaccounted for confounders.* 

## 9 Reproducibility

```{r session-info, cache = FALSE}
sessionInfo()
```
